---
title: "NFL Home Field Advantage Analysis"
author: "Daniel Carney, Mohammed Alghamdi, Siddhant Bhagat"
date: today
format:
  pdf:
    toc: true
    number-sections: true
    fig-pos: "H"
    fig-width: 7
    fig-height: 4.5
execute:
  echo: false
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: load packages
#| include: false

## loading the tidyverse for all data manipulation and visualization
library(tidyverse)
library(knitr)
```

# Introduction

## Research Question

Is there any statistical truth behind home field advantage in the NFL?

We have access to data on the outcome of every NFL game from 2018
through the 2023 season. Using this data we will investigate whether
home teams win at a rate that is meaningfully higher than 50%, and
whether the points they score relative to away teams supports the same
conclusion. Our approach is confirmatory data analysis (CDA). We begin
with a specific directional hypothesis that home teams win more than 50%
of games and use a formal statistical test (a proportion test) to
evaluate the evidence for that claim. This distinguishes our work from
exploratory analysis, where hypotheses emerge from the data rather than
preceding it.

##Approach

All of our data analysis work follows the Tidyverse paradigm rather than
Base R. This means we use dplyr for data manipulation, ggplot2 for
visualisation, and the pipe operator for composition. We chose the
Tidyverse paradigm because it produces readable, consistent, and
reproducible code. Each step in a pipeline makes its intent explicit,
which supports collaboration across team members and makes the Plan Code
Improve Polish (PCIP) process easier to follow. The Tidyverse also
integrates naturally with Quarto (QMD), which is how we render our final
output.

# Data

## Data Source and Provenance

Our data comes from the SCORE Sports Data Repository, a curated academic
sports data resource. The specific dataset is:

(SCORE Network, n.d.)

The dataset contains one row per NFL game, covering every regular season
and playoff game from 2018 through the end of the 2023 season.

## FAIR and CARE Principles

**FAIR:** The SCORE repository makes this data findable via a stable
URL, accessible at no cost, interoperable as a CSV file that any
statistical software can read, and reusable under an open academic
licence with a clear citation path. Our analysis satisfies these
principles because we cite the source formally, import the data via a
coded command rather than a manual download, and store both the raw data
and our code in our GitHub repository so the work can be reproduced.

**CARE :** This dataset contains only aggregate game level outcomes with
no personally identifiable information about individual players beyond
team level identifiers. There are no privacy concerns. The data was
collected and published by an academic institution for the purpose of
supporting student learning, which aligns with the principle of
Collective Benefit. We take Responsibility for using it appropriately by
not overgeneralising beyond the six season window it covers.

## Cases and Case Attributes

In the raw data, a **case** is a single NFL game. Each game is one
observation, that is, one row in the dataset. The attributes (variables)
recorded for each game are:

| Attribute      | Type      | Description                               |
|----------------|-----------|-------------------------------------------|
| `season`       | Integer   | The NFL season year (2018 to 2023)        |
| `game_id`      | Character | Unique game identifier                    |
| `week`         | Integer   | Week number within the season             |
| `home_team`    | Character | Three letter team abbreviation            |
| `away_team`    | Character | Three letter team abbreviation            |
| `home_score`   | Integer   | Points scored by the home team (points)   |
| `away_score`   | Integer   | Points scored by the away team (points)   |
| `game_outcome` | Numeric   | 1.0 = home win, 0.5 = tie, 0.0 = away win |
| `score_diff`   | Integer   | home score minus away score (points)      |

After wrangling, we reshape to one row per team per game so each game
produces two cases: one representing the home team perspective and one
representing the away team perspective. In this reshaped data a **case**
is a single team game appearance. The additional derived attributes we
create are `team`, `opponent`, `points_for` (points), `points_against`
(points), `location` (Home or Away), `result` (Win or Loss), and
`total_points` (points).

# Data Preparation

```{r}
#| label: cleaning

# Cleaning the NFL data
## The goal is to clean the NFL data to be in a glyph ready form that can be
## used to create visualizations in order for us to make insights about whether
## or not home field advantage is a real phenomenon in the NFL

# Step 1: Load Packages
## Needed Packages: tidyverse
library(tidyverse)

# Step 2: Load Data
## Needed data: NFL DATA
nfl_raw <- read_csv("nfl_mahomes_era_games.csv", show_col_types = FALSE)

## Step 3: filter the raw data
nfl_filtered <- nfl_raw %>%
  filter(game_type != "SB") %>%   #remove super bowls
  filter(score_diff != 0)         #remove the tie games

## Step 4: Separate from one row per game to one row per team
#this is important for comparing a teams home vs away performance

#first make a total score column to work with
nfl_vars <- nfl_filtered %>%
  mutate(total_points = home_score + away_score)

#Home team rows
home_teams <- nfl_vars %>%
  mutate(
    team = home_team,
    opponent = away_team,
    points_for = home_score,
    points_against = away_score,
    location = "Home",
    result = ifelse(game_outcome == 1, "Win", "Loss")
  )

#Away team rows
away_teams <- nfl_vars %>%
  mutate(
    team = away_team,
    opponent = home_team,
    points_for = away_score,
    points_against = home_score,
    location = "Away",
    result = ifelse(game_outcome == 0, "Win", "Loss")
  )

## Step 5 Combine & Arrange

#orders the tables so you can view them by team.
nfl_clean <- bind_rows(home_teams, away_teams) %>%
  arrange(team, season, week) %>%
  select(season, game_type, week, team, opponent, location, points_for,
         points_against, result, total_points, game_id)

## Step 6 Ensure data is categorized correctly

#this step is what allows for analysis because the groups are treated as
#categories and not mathematical numbers.
nfl_clean <- nfl_clean %>%
  mutate(
    season = as.factor(season),
    result = as.factor(result),
    location = as.factor(location)
  )
```

We removed `r nrow(nfl_raw) - nrow(nfl_filtered)` games from the raw
dataset: 6 Super Bowls (neutral site games where home and away labels
are not meaningful) and `r sum(nfl_raw$score_diff == 0)` tied games
(because our analysis uses a binary win/loss outcome). After filtering
we have `r nrow(nfl_filtered)` games, which produce `r nrow(nfl_clean)`
team game rows in the reshaped dataset.

# Descriptive Statistics

```{r}
#| label: processing

# Processing the NFL data
## The goal is to take the cleaned NFL data and summarise home win rate by
## season. This chunk does the calculation and saves the result so the
## presentation section can load it and make the visual.

## Step 3: Summarise home win rate by season
## filtering to home rows only so each game counts once
season_summary <- nfl_clean %>%
  filter(location == "Home") %>%
  group_by(season) %>%
  summarise(
    games    = n(),
    win_rate = mean(result == "Win"),
    .groups  = "drop"
  )

## Step 4: Saving the result
write_csv(season_summary, "season_summary.csv")
```

```{r}
#| label: summary table

## home vs away summary using dplyr::summarise and knitr::kable
location_stats <- nfl_clean %>%
  group_by(location) %>%
  summarise(
    Games           = n(),
    Wins            = sum(result == "Win"),
    `Win Rate`      = round(mean(result == "Win"), 4),
    `Avg Points For`     = round(mean(points_for), 2),
    `Avg Points Against` = round(mean(points_against), 2),
    `Avg Margin`         = round(mean(points_for - points_against), 2),
    .groups = "drop"
  )

kable(
  location_stats,
  caption  = "Summary statistics: home vs. away team game appearances (2018 to 2023)",
  booktabs = TRUE,
  align    = "lrrrrrr"
)
```

```{r}
#| label: season table

## season by season breakdown of home performance
season_table <- nfl_clean %>%
  filter(location == "Home") %>%
  group_by(season) %>%
  summarise(
    Games      = n(),
    Wins       = sum(result == "Win"),
    `Win Rate` = round(mean(result == "Win"), 3),
    `Avg Home Points` = round(mean(points_for), 2),
    `Avg Away Points` = round(mean(points_against), 2),
    `Avg Margin`      = round(mean(points_for - points_against), 2),
    .groups = "drop"
  )

kable(
  season_table,
  caption  = "Home team performance by season",
  booktabs = TRUE,
  align    = "lrrrrrr"
)
```

```{r}
#| label: proportion test

## formal hypothesis test
## H0: home win rate = 50% (no home advantage)
## H1: home win rate != 50% (home advantage exists)
home_only  <- nfl_clean %>% filter(location == "Home")
home_games <- nrow(home_only)
home_wins  <- sum(home_only$result == "Win")

prop_result <- prop.test(x = home_wins, n = home_games, p = 0.5)
```

Across all `r home_games` home team appearances in the dataset, home
teams won `r home_wins` games for an overall win rate of
**`r round(100 * home_wins / home_games, 1)`%**. Home teams also
outscored their opponents by an average of **+1.57 points per game**
(total margin of 2,586 points across 1,644 games). A proportion test
against the 50% no advantage null hypothesis yields an exact p value of
**0.000286**, which is strong evidence that the home win rate is not due
to chance.

# Visualisations and Analysis

## Home and Away Win Rate (Daniel)

```{r}
#| label: fig pie home
#| fig-cap: "Home team win rate across the 2018 to 2023 NFL seasons. Green represents wins and red represents losses."
#| fig-alt: "A pie chart showing home team results. The green win segment is larger than the red loss segment, representing a 54.5% win rate for home teams across 1644 games."

## Pie chart contribution by Daniel Carney
## Group: Daniel Carney, Mohammed Alghamdi, Siddhant Bhagat

## Step 7 Create a visualization for home team win rate

#Count the home wins and losses
home_result_counts <- home_teams %>%
  count(result)

ggplot(home_result_counts, aes(x = " ", y = n, fill = result)) +
  #ggplot does not have a native pie chart so I used column(bar chart) and polar
  #in order to create a pie chart out of bar charts
  geom_col(width = 1, color = "white") +
  coord_polar("y", start = 0) +
  scale_fill_manual(
    values = c("Loss" = "#FF0000", "Win" = "#008000"),
    labels = c("Loss", "Win")
  ) +
  labs(
    title = "Home Team Win Rate",
    fill  = "Result"
  )
```

```{r}
#| label: fig pie away
#| fig-cap: "Away team win rate across the 2018 to 2023 NFL seasons. Green represents wins and red represents losses."
#| fig-alt: "A pie chart showing away team results. The red loss segment is larger than the green win segment, representing a 45.5% win rate for away teams across 1644 games."

## Pie chart contribution by Daniel Carney
## Group: Daniel Carney, Mohammed Alghamdi, Siddhant Bhagat

## Step 8 Create a visualization for away team win rate
#Count the away wins and losses
away_result_counts <- away_teams %>%
  count(result)

ggplot(away_result_counts, aes(x = " ", y = n, fill = result)) +
  #ggplot does not have a native pie chart so I used column(bar chart) and polar
  #in order to create a pie chart out of bar charts
  geom_col(width = 1, color = "white") +
  coord_polar("y", start = 0) +
  scale_fill_manual(
    values = c("Loss" = "#FF0000", "Win" = "#008000"),
    labels = c("Loss", "Win")
  ) +
  labs(
    title = "Away Team Win Rate",
    fill  = "Result"
  )
```

The figure the win and loss split for home and away teams
respectively.,confirming that home teams win more often than away teams.
Home teams won 54.5% of games while away teams won only 45.5%, a gap of
9 percentage points. This visual pattern is consistent with the
proportion test result (p = 0.000286), confirming the gap is
statistically significant and not driven by random variation.

## Home Win Rate by Season (Siddhant)

```{r}
#| label: fig season line
#| fig-cap: "Home team win rate by NFL season from 2018 to 2023. The dashed line marks the 50% no advantage baseline. The 2020 season, played largely without fans due to COVID restrictions, is the only year the home win rate fell below 50%."
#| fig-alt: "A line chart with red points and lines showing home win rate per season. Values start at 59.8% in 2018, drop to 49.8% in 2020, then recover to 56.3% by 2023. A dashed black line sits at the 50% mark."

## Line chart contribution by Siddhant Bhagat
## Group: Daniel Carney, Mohammed Alghamdi, Siddhant Bhagat

## Presenting the NFL home advantage analysis
## The goal is to take the saved season summary and produce a line chart of
## home win rate by season.

## Step 3: Plot home win rate by season
## line drops to ~50% in 2020 then recovers which supports the
## idea that crowd noise is part of what drives home field advantage since
## we see a fluctuation in win rate > 5%

season_summary %>%
  ggplot(aes(x = season, y = win_rate, group = 1)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "black") +
  geom_line(color = "red", linewidth = 1) +
  geom_point(size = 3, color = "red") +
  geom_text(aes(label = scales::percent(win_rate, accuracy = 0.1)),
            vjust = -1, size = 3.5) +
  scale_y_continuous(labels = scales::percent_format(),
                     limits = c(0.4, 0.7)) +
  labs(
    title    = "Home win rate by season",
    subtitle = "Dashed line = 50% (no advantage)",
    x        = "Season",
    y        = "Home win rate"
  ) +
  theme_minimal(base_size = 12)
```

The win% line sits above 50% in every season except 2020, where the win
rate dropped to 49.8%, effectively making it a coin flip. The 2020 NFL
season was played largely without fans due to COVID restrictions, which
distinguishes it from every other season in our window. The fact that
home advantage essentially disappeared when crowds were absent, and then
recovered immediately once fans returned in 2021 (51.9%), climbing
further to 56.9% by 2022, strongly implicates crowd noise and home
atmosphere as a driver of the effect rather than travel fatigue or
scheduling alone. This pattern provides a natural quasi experiment
embedded in our data that goes beyond what a simple overall average
could reveal.

## Home vs Away Average Points by Season (Mohammed)

```{r}
#| label: fig bar avg points season
#| fig-cap: "Average points scored per game by home teams versus away teams across each NFL season from 2018 to 2023. The 2020 season (no fans) shows a notably narrower gap than the surrounding seasons."
#| fig-alt: "A grouped bar chart with two bars per season. In every season except 2020, the green Home bar is taller than the red Away bar. Each bar is labelled with its rounded value. The gap between Home and Away is widest in 2018 and 2022 and narrowest in 2020."

## Bar graph contribution by Mohammed Alghamdi
## Group: Daniel Carney, Mohammed Alghamdi, Siddhant Bhagat

# PLAN: visualize how the average points scored per game by home teams
# compares to that of away teams across each season in our window. If
# home field advantage drives scoring as well as win rate, we expect
# the home bar to sit above the away bar in most seasons, with a
# possible narrowing in 2020 (no fans) consistent with the line chart.

# CODE: starting from the per-game filtered data, summarise each
# season's average home and away score, then reshape to long form so
# location is a single column we can map to fill in ggplot.
avg_points_by_season <- nfl_filtered %>%
  group_by(season) %>%
  summarise(
    Home = mean(home_score),
    Away = mean(away_score),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols      = c(Home, Away),
    names_to  = "location",
    values_to = "avg_pts"
  )

# IMPROVE: factor levels so Home plots first within each season group.
avg_points_by_season <- avg_points_by_season %>%
  mutate(location = factor(location, levels = c("Home", "Away")))

# POLISH: dodge bars side by side per season, label each bar with its
# rounded value, use the same green/red palette as the pie charts, and
# give axes units in points and seasons.
ggplot(avg_points_by_season,
       aes(x = season, y = avg_pts, fill = location)) +
  geom_col(position = position_dodge(width = 0.9), width = 0.8) +
  geom_text(
    aes(label = round(avg_pts, 1)),
    position = position_dodge(width = 0.9),
    vjust    = -0.5,
    size     = 3
  ) +
  scale_fill_manual(values = c("Home" = "#008000",
                               "Away" = "#FF0000")) +
  labs(
    title = "Home vs Away Average Points by Season",
    x     = "Season (year)",
    y     = "Average points per game (points)",
    fill  = "Team location"
  ) +
  theme_minimal(base_size = 12)
```

There are two bars per year between 2018 and 2023, one green for the
home team and another red for the away team. There is hardly any year
where the green bar is smaller than the red bar, except for 2020 when
there is hardly any difference between the two bars. The creation of
the chart involved grouping the data based on seasons, computing mean
values for both home and away teams, converting the data into long
format, and creating a bar chart using ggplot. Home teams had a higher
score compared to away teams for most years. In 2020, however, the
score for the home team was nearly equal to that of the away team.

# Conclusion

## Key Finding

There is strong statistical evidence for home field advantage in the NFL
during the 2018 to 2023 Mahomes era. Home teams won 54.5% of games, a
rate significantly above 50% (p = 0.000286), and outscored away teams by
+1.57 points per game on average. The effect is consistent across most
seasons and holds for both win rate and scoring margin.

## The 2020 Natural Experiment

The most compelling insight from this analysis is the 2020 season. When
fans were absent the home win rate dropped to 49.8% and the scoring gap
between home and away teams narrowed. When fans returned in 2021 the
advantage recovered and by 2022 it had climbed to 56.9%. This pattern
suggests that crowd noise and home atmosphere are meaningful
contributors to the advantage rather than structural factors like travel
distance or scheduling alone.

## Limitations

This analysis covers six seasons and is therefore specific to this era
of the NFL. The 2020 fan restriction was not uniform as some stadiums
admitted limited fans later in the season, which means the 2020 data
point is an approximation of a fully fanless season rather than a clean
treatment. Future work could examine team level variation in home
advantage or compare across sports where crowds may have different
effects.

# Author Contributions

**Daniel Carney** : led the data cleaning and reshaping process,
designing the pipeline to transform the raw game level data into a team
game format suitable for home versus away comparisons.

**Mohammed Alghamdi :** The author of the README file and contributor
to the bar graph showing the average points earned by teams playing
at home versus those playing away during 2018–2023. Also checked that
the code followed rubric requirements.

**Siddhant Bhagat :** was responsible for the statistical processing and
inferences, created plot for season to winrate performance on a year by
year basis and made inferences from the resulting data

# References

Yurko, R.Yurko, R. (2024, July 8). *National Football League game
outcomes* \[Data set\]. SCORE Sports Data Repository.
https://data.scorenetwork.org/football/nfl-game-outcomes.html

# Code Appendix

```{r}
#| label: appendix cleaning
#| echo: true
#| eval: false

# Cleaning the NFL data
## Brief description of what your goals are
## The goal is to clean the NFL data to be in a glyph ready form that can be
## used to create visualizations in order for us to make insights about whether
## or not home field advantage is a real phenomenon in the NFL

# Step 1: Load Packages
## Needed Packages: tidyverse
library(tidyverse)

# Step 2: Load Data
## Needed data: NFL DATA
nfl_raw <- read_csv("nfl_mahomes_era_games.csv")

## Step 3: filter the raw data
nfl_filtered <- nfl_raw %>%
  filter(game_type != "SB") %>%   #remove super bowls
  filter(score_diff != 0)         #remove the tie games

## Step 4: Separate from one row per game to one row per team
#this is important for comparing a teams home vs away performance

#first make a total score column to work with
nfl_vars <- nfl_filtered %>%
  mutate(total_points = home_score + away_score)

#Home team rows
home_teams <- nfl_vars %>%
  mutate(
    team = home_team,
    opponent = away_team,
    points_for = home_score,
    points_against = away_score,
    location = "Home",
    result = ifelse(game_outcome == 1, "Win", "Loss")
  )

#Away team rows
away_teams <- nfl_vars %>%
  mutate(
    team = away_team,
    opponent = home_team,
    points_for = away_score,
    points_against = home_score,
    location = "Away",
    result = ifelse(game_outcome == 0, "Win", "Loss")
  )

## Step 5 Combine & Arrange

#orders the tables so you can view them by team.
nfl_clean <- bind_rows(home_teams, away_teams) %>%
  arrange(team, season, week) %>%
  select(season, game_type, week, team, opponent, location, points_for,
         points_against, result, total_points, game_id)

## Step 6 Ensure data is categorized correctly

#this step is what allows for analysis because the groups are treated as
#categories and not mathematical numbers.
nfl_clean <- nfl_clean %>%
  mutate(
    season = as.factor(season),
    result = as.factor(result),
    location = as.factor(location)
  )
```

```{r}
#| label: appendix processing
#| echo: true
#| eval: false

# Processing the NFL data
## Brief description of what your goals are
## The goal is to take the cleaned NFL data and summarise home win rate by
## season. This file does the calculation and saves the result so the
## presentation file can load it and make the visual.

# Step 1: Load Packages
library(tidyverse)

# Step 2: Load the cleaned data
## Run the cleaning script to bring nfl_clean into the environment
source("Cleaning_Data_DC.R")

## Step 3: Summarise home win rate by season
## filtering to home rows only so each game counts once
season_summary <- nfl_clean %>%
  filter(location == "Home") %>%
  group_by(season) %>%
  summarise(
    games    = n(),
    win_rate = mean(result == "Win"),
    .groups  = "drop"
  )

## Step 4: Saving the result
write_csv(season_summary, "season_summary.csv")
```

```{r}
#| label: appendix visualisation
#| echo: true
#| eval: false

# Home Team Win Visual
## Brief description of what your goals are
## The goal is to clean the NFL data to be in a glyph ready form that can be
## used to create visualizations in order for us to make insights about whether
## or not home field advantage is a real phenomenon in the NFL

## Step 7 Create a visualization for home team win rate

#Count the home wins and losses
home_result_counts <- home_teams %>%
  count(result)

ggplot(home_result_counts, aes(x = " ", y = n, fill = result)) +
  #ggplot does not have a native pie chart so I used column(bar chart) and polar
  #in order to create a pie chart out of bar charts
  geom_col(width = 1, color = "white") +
  coord_polar("y", start = 0) +
  scale_fill_manual(
    values = c("Loss" = "#FF0000", "Win" = "#008000"),
    labels = c("Loss", "Win")
  ) +
  labs(
    title = "Home Team Win Rate",
    fill  = "Result"
  )

## Step 8 Create a visualization for away team win rate
#Count the away wins and losses
away_result_counts <- away_teams %>%
  count(result)

ggplot(away_result_counts, aes(x = " ", y = n, fill = result)) +
  #ggplot does not have a native pie chart so I used column(bar chart) and polar
  #in order to create a pie chart out of bar charts
  geom_col(width = 1, color = "white") +
  coord_polar("y", start = 0) +
  scale_fill_manual(
    values = c("Loss" = "#FF0000", "Win" = "#008000"),
    labels = c("Loss", "Win")
  ) +
  labs(
    title = "Away Team Win Rate",
    fill  = "Result"
  )
```

```{r}
#| label: appendix presentation
#| echo: true
#| eval: false

# Presenting the NFL home advantage analysis
## Brief description of what your goals are
## The goal is to take the saved season summary and produce a line chart of
## home win rate by season.

# Step 1: Load Packages
library(tidyverse)

# Step 2: Load the saved result
season_summary <- read_csv("season_summary.csv")

## Step 3: Plot home win rate by season
## line drops to ~50% in 2020 then recovers which supports the
## idea that crowd noise is part of what drives home field advantage since
## we see a fluctuation in win rate > 5%

season_summary %>%
  ggplot(aes(x = season, y = win_rate, group = 1)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "black") +
  geom_line(color = "red", linewidth = 1) +
  geom_point(size = 3, color = "red") +
  geom_text(aes(label = scales::percent(win_rate, accuracy = 0.1)),
            vjust = -1, size = 3.5) +
  scale_y_continuous(labels = scales::percent_format(),
                     limits = c(0.4, 0.7)) +
  labs(
    title    = "Home win rate by season",
    subtitle = "Dashed line = 50% (no advantage)",
    x        = "Season",
    y        = "Home win rate"
  ) +
  theme_minimal(base_size = 12)
```

```{r}
#| label: appendix bar avg points season
#| echo: true
#| eval: false

# Bar graph: Home vs Away Average Points by Season
## Author: Mohammed Alghamdi
## Group: Daniel Carney, Mohammed Alghamdi, Siddhant Bhagat
## Goal: compare the average points scored per game by home teams and
## away teams across each season from 2018 to 2023. This complements the
## pie charts (overall win rate) and the line chart (home win rate over
## time) by showing the scoring side of the same effect.

# Step 1: Load Packages
library(tidyverse)

# Step 2: Use the filtered per-game data from the cleaning chunk
## (nfl_filtered is already in the environment from Cleaning_Data_DC.R)

# PLAN: per season, summarise mean home_score and mean away_score, then
# pivot to long format so location is a single column for ggplot fill.

# CODE
avg_points_by_season <- nfl_filtered %>%
  group_by(season) %>%
  summarise(
    Home = mean(home_score),
    Away = mean(away_score),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols      = c(Home, Away),
    names_to  = "location",
    values_to = "avg_pts"
  )

# IMPROVE: factor levels so Home plots first within each season group.
avg_points_by_season <- avg_points_by_season %>%
  mutate(location = factor(location, levels = c("Home", "Away")))

# POLISH: dodged bars per season, value labels above each bar, palette
# matched to the pie charts, axis units called out.
ggplot(avg_points_by_season,
       aes(x = season, y = avg_pts, fill = location)) +
  geom_col(position = position_dodge(width = 0.9), width = 0.8) +
  geom_text(
    aes(label = round(avg_pts, 1)),
    position = position_dodge(width = 0.9),
    vjust    = -0.5,
    size     = 3
  ) +
  scale_fill_manual(values = c("Home" = "#008000",
                               "Away" = "#FF0000")) +
  labs(
    title = "Home vs Away Average Points by Season",
    x     = "Season (year)",
    y     = "Average points per game (points)",
    fill  = "Team location"
  ) +
  theme_minimal(base_size = 12)
```
